### 2018-07-08 report.tex
Extended chapter 2

%%  Begin title page
%%  ==================================================================
    \thispagestyle{empty}
    \ULCornerWallPaper{1}{nat-farve.pdf}
    \ULCornerWallPaper{1}{nat-en.pdf}
    \begin{adjustwidth}{-3cm}{-1.5cm}
    \vspace*{-1cm}
    \textbf{\Huge Bioinformatics Master Thesis} \\
    \vspace*{2.5cm} \\
    \textbf{\Huge Topics in Mass Sectrometry and Proteomics} \\
    \vspace*{.1cm} \\
    %{\huge No subtitle is available} \\
    \begin{tabbing}
    % adjust the hspace below for the longest author name
    Antonio Ortega Jiménez \hspace{1cm} \= \texttt{<ntoniohu@gmail.com>}
    %Second Author \> \texttt{<second@author.com>} \\       
    \\[10.5cm]

@@

In the label-free quantification approach, peptides from different samples are not labelled differently and are thus distinguished by their presence in different, independent \ac{MS} runs. In order to account for inter-run variability in peptide identifications and \ac{RT}, a match-between-runs (MBR) processing can be carried out for pair-wise transfer of peptide identities from runs where identification was successful to those were it wasn't due to the randomness in the identification process. The transfer can be made based on shared precursor mass and corrected retention times CITE MOFF.

@@

\section*{Summary}
A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI, PeptideShaker, moFF and MSqRob CITE THEM, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software. The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) feature extraction, (IV) protein inference and (V) relative quantification. Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.  A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}. The results exhibited an accuracy similar to those achieved by the MaxLFQ \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.

@@

The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}. Briefly, the \textit{Homo sapiens} and \textit{E. coli (strain K12)} proteomes were mixed in 1:1 and 1:3 proportions, with 3 replicates for each combination. Moreover, each of the 3 replicates of the 2 conditions was analysed over 24 fractions. This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files. The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format).

@@

The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}. By default, the FDR was set to 1\%. Output was extracted via the Default PSM report txt file, available in the \texttt{ReportCLI} utility.

@@

@@

The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra. PeptideShaker quality control and filtering capabilities (see figures \ref{figure:qc_validation}) carried out the required search results validation. As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values. Likewise, the m/z error was found to be closer to 0 on validated

@@

\begin{figure}[!h]
\centering
\begin{subfigure}{.4\textwidth}
  \centering
  \caption*{A}
  \includegraphics[width=.9\linewidth]{target_vs_decoy}
\end{subfigure}
\begin{subfigure}{.4\textwidth}
  \centering
    \caption*{B}
  \includegraphics[width=.9\linewidth]{score_vs_confidence}
\end{subfigure}
\bigskip

\begin{subfigure}{.4\textwidth}
  \centering
    \caption*{C}
  \includegraphics[width=.8\linewidth]{qc}
\end{subfigure}
\begin{subfigure}{.4\textwidth}
  \centering
    \caption*{D}
  \includegraphics[width=.9\linewidth]{roc}
\end{subfigure}
\caption{Quality control and validation of the 13th fraction of the third replicate in condition L. \textbf{A} Score distribution for matches to the decoy and the target databases. \textbf{B} Evolution of the PSM score with confidence. The implemented cutoff at FDR of 1 \% is displayed with a red vertical line. \textbf{C} Distribution of the difference between the predicted and measured \ac{m/z} values, segregated by validation status. \textbf{D} ROC curve built upon the number of false positives and negatives estimated from the decoy search. The cutoff is displayed as a red dot.}
\label{figure:qc_validation}
\end{figure}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{match_percent}
\caption{Percentage of matched spectra in all samples. The total number of spectra per sample ranged between 5894 and 20249. \textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa. \textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation.}
\label{fig:match_percent}
\end{figure}

@@

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{match_percent}
\caption{}
\label{fig:match_percent}
\end{figure}

@@

>>>>

%%  Begin title page
%%  ==================================================================
    \thispagestyle{empty}
    \ULCornerWallPaper{1}{nat-farve.pdf}
    \ULCornerWallPaper{1}{nat-en.pdf}
    \begin{adjustwidth}{-3cm}{-1.5cm}
    \vspace*{-1cm}
    \textbf{\Huge Bioinformatics Master Thesis} \\
    \vspace*{2.5cm} \\
    \textbf{\Huge Development of a label-free quantification proteomics pipeline} \\
    \vspace*{.1cm} \\
    %{\huge No subtitle is available} \\
    \begin{tabbing}
    % adjust the hspace below for the longest author name
    Antonio Ortega Jiménez \hspace{1cm} \= \texttt{<ntoniohu@gmail.com>}
    %Second Author \> \texttt{<second@author.com>} \\       
    \\[10.5cm]

@@

In the label-free quantification approach, peptides from different samples are not labelled differently and are thus distinguished by their presence in different, independent \ac{MS} runs. In order to account for inter-run variability in peptide identifications and \ac{RT}, a match-between-runs (MBR) processing can be carried out for pair-wise transfer of peptide identities from runs where identification was successful to those were it wasn't due to the randomness in the identification process. The transfer can be made based on shared precursor mass and corrected retention times CITE MOFF. Furthermore, in order to collect consistent \ac{XIC}, a feature extraction step can be  executed to extract the apex intensity of the identified peak clusters \ref{figure:moff}.

\begin{figure}[!h]
\centering
\begin{subfigure}{.45\textwidth}
    \caption*{A}
\includegraphics[width=0.9\linewidth]{moff_workflow}
\end{subfigure}
\begin{subfigure}{.45\textwidth}
    \caption*{B}
\includegraphics[width=0.9\linewidth]{apex_intensity}
\end{subfigure}
\caption{Illustration of the \ac{MBR} and apex intensity extraction steps. \textbf{A} the information gathered from matches in replicate runs is collected for a reanalysis of the spectra. \textbf{B} A retention time window can be used to screen the peak clusters and highest \ac{MS1} intensity of a peak cluster (apex intensity). Taken from CITE MOFF}
\label{figure:moff}
\end{figure}

@@

\section*{Summary}
A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI, PeptideShaker, moFF and MSqRob CITE THEM, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software. The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) MBR and feature extraction, (IV) protein inference and (V) relative quantification. Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.  A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}. The results exhibited an accuracy similar to those achieved by the MaxLFQ \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.

@@

The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}. Briefly, the \textit{Homo sapiens} and \textit{E. coli (strain K12)} proteomes were mixed in 1:1 (condition L) and 1:3 (condition H) proportions, with 3 replicates for each combination. Moreover, each of the 3 replicates of the 2 conditions was analysed over 24 fractions. This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files. One file was missing in the repository. The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format).

@@

The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}. By default, the FDR was set to 1\%. PEP and confidence statistics were computed using the PeptideShaker built-in algorithms. Output was extracted via the Default PSM report txt file, available in the \texttt{ReportCLI} utility.

@@

\subsection{Code implementation}

@@

The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra. PeptideShaker quality control and filtering capabilities (see figure \ref{figure:qc_validation}) carried out the required search results validation. As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values. Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1 \% FDR filter. The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.

\begin{figure}[!h]
\centering
\begin{subfigure}{.45\textwidth}
  \centering
  \caption*{A}
  \includegraphics[width=.9\linewidth]{target_vs_decoy}
\end{subfigure}
\begin{subfigure}{.45\textwidth}
  \centering
    \caption*{B}
  \includegraphics[width=.9\linewidth]{score_vs_confidence}
\end{subfigure}
\bigskip

\begin{subfigure}{.45\textwidth}
  \centering
    \caption*{C}
  \includegraphics[width=.8\linewidth]{qc}
\end{subfigure}
\begin{subfigure}{.45\textwidth}
  \centering
    \caption*{D}
  \includegraphics[width=.9\linewidth]{roc}
\end{subfigure}
\caption{Quality control and validation of the 13th fraction of the third replicate in condition L. \textbf{A} Score distribution for matches to the decoy and the target databases. \textbf{B} Evolution of the PSM score with confidence. The implemented cutoff at FDR of 1 \% is displayed with a red vertical line. \textbf{C} Distribution of the difference between the predicted and measured \ac{m/z} values, segregated by validation status. \textbf{D} ROC curve built upon the number of false positives and negatives estimated from the decoy search. The cutoff is displayed as a red dot.}
\label{figure:qc_validation}
\end{figure}

The PSM confidence is defined as $1-PEP$, where PEP stands for the Posterior Error Probability. Also known as local FDR, the PEP is an estimate of the probability of a given PSM of being an incorrect assignment. Thus, the confidence is the probability of the PSM being a correct assigment CITE NEVINZKY. The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence of at least 65\%.

@@

\begin{figure}[!h]
\includegraphics[width=0.8\textwidth]{match_percent}
\caption{Percentage of matched spectra in all samples. The total number of spectra per sample ranged between 5894 and 20249. \textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa. \textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation. The sixth fraction of the first replicate in condition H was missing in the PRIDE data repository.}
\label{fig:match_percent}
\end{figure}

\newpage

@@

The match between runs step allows for increased identifications by transferring successful matches between replicate runs. The results of this process for the 13th fraction of the L condition is shown in figures \ref{fig:mbr_summary} and \ref{fig:mbr})

\begin{figure}[!h]
\centering
\includegraphics[width=0.5\textwidth]{mbr_summary}
\caption{Count of identifications on each run segreagated by source. More than 4k spectra were identified and validated by SearchGUI+peptideShaker. The MBR acts as an imputation step where extra identifications are performed by gathering the information collected from identifications in other samples and reprocessing the spectra files. As a consequence, more identifications with a lesser fraction of missing datapoints are achieved. In this particular case, hundreds of identifications were accomplished}

\label{fig:mbr_summary}
\end{figure}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{mbr}
\caption{Match Between Runs with moFF. Every dot represents a peptide shared across 2 runs. The coordinate system illustrates its retention time on the first run on the x axis and the difference with the second run on the y axis. The color depicts whether the identification was carried out during the PSM process, or thanks to a cross-identification achieved by the MBR module.}
\label{fig:mbr}
\end{figure}

Once as many identifications as possible were gathered, a refinement of the measured MS1 intensity can be implemented to select the apex of every peak cluster, which yields robust intensity measurements for each sample (see figure \ref{fig:apex_intensity}). The extracted apex intensity can be used as a proxy for peptide abundance in the following quantification step.

\begin{figure}[!h]
\begin{subfigure}{.9\textwidth}
  \centering
    \caption*{A}
\includegraphics[width=.9\linewidth]{peptide_profile}
\end{subfigure}
\bigskip

\begin{subfigure}{.9\textwidth}
  \centering
    \caption*{B}
\includegraphics[width=.9\linewidth]{density_ratio}
\end{subfigure}
\caption{\textbf{A} The apex intensity profile across fractions for 2 different peptides, one from \textit{Homo sapiens}, and one from \textit{E. coli}. The figure illustrates the intrinsic intensity variability between technical replicates, particularly in the case of the \textit{E. coli} QLLMMSGFDR peptide, as it was almost non-existent in one of the runs. \textbf{B} The expected pattern of overall similar intensities for the \textit{Homo sapiens} data and 3-fold higher intensities for the \textit{E. coli} data in condition H was observed, confirming the good performance of the protocol.}
\label{fig:apex_intensity}
\end{figure}

@@

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{graph}
\end{figure}
